Demography and linked selection in wild and domesticated maize

### 2015-09-09 manuscript.tex
ok. good with methods and results and significance and all but last section of discussion. last section of discussion, abstract, and intro need work.

\subsection{Hard sweeps do not shape genome-wide diversity in maize}

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation as the primary source of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  
Assuming a mutation rate of $3 \times 10^{-8}$ \citep{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \citep{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \citep[e.g.][]{studer2011}. 
Indeed, many adaptive traits in both maize \citex and teosinte \citex are highly quantitative \jri{cite wallace for maize? weber and/or lauter for teosinte?} and adaptation in both maize \citep{hufford2012} and teosinte \citep{pyhajarvi2013complex} has involved selection on regulatory variation.  \jri{if number is nice, add something like "\X of the windows of top H12 score genome-wide occur outside of genes"}

\jri{paragraph needs work. maybe eliminate/move? i do think we need some comparison to other stuff, but the broader Ne-selection efficiency argument should come later.}
The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this explanation into question.}

\subsection{Demography of domestication}
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites.  
By utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without requiring assumptions about its duration.

One surprising result from our model is the estimation of the timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological \cite{piperno2009starch} estimates, we instead argue this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may also be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

The estimated bottleneck of $\approx 5\%$ of the ancestral teosinte population seems low given that maize landraces exhibit $\approx 80\%$ of the diversity of teosinte \citep{hufford2012}, but our model suggests that the effects of the bottleneck on diversity are likely ameliorated to some extent by both gene flow and rapid post-domestication growth (Figure \ref{fig:bottleneck}).   
Although we estimate that the modern effective size of maize is larger than teosinte, the small size of our sample reduces our power to identify the low frequency alleles most sensitive to rapid population growth \cite{keinan2012}.  
A better estimate comes from our analysis of singleton diversity in a large sample of maize landraces \cite{Hearne2015}, which suggests a modern effective size of $\approx 1$ million, or nearly eight times higher than modern teosinte.
Even this estimate, however, is likely downwardly biased due to missing data, sample size, and ascertainment in the genotyping data.
An alternative approach is to use estimates of the census size of modern maize: there are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, likely planted at a density of $\approx 25,000$ individuals  per hectare \cite{baden2001culture}.
Assuming the effective size is only $\approx 0.4\%$ of the census size (i.e. 1 ear for every 1000 plants), this still implies a modern effective population size of more than four billion.
While these rough estimates are likely poor, all of the evidence points to the fact that the effective size of modern maize is extremely large.

@@

\jri{I would like to see a supplemental figure or maybe just number here on the partial correlation in both teosinte and maize of diversity with recombination after accounting for gene density.  do this in 500kb windows and then in 10kb windows.  we should do the correlation with both pi and singletons.  This let's us compare more explictly with studies like C-D who evaluate overall patterns of linked selection. will be especially cool if we don't see a big difference in big windows but we do in 10kb windows.  }
A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. \jri{inbreds stuff OK, but should be shortened.  and we need more/broader. Chia, Rogers-Melnick, Mezmouk all looked at load. does this have implications for large effect load (i.e. should bneck have purged those?), for heterosis, architecture of quant. traits, etc.}

@@

\begin{materials}

@@

\subsection{Plant materials}
We made use of published sequences from inbred accessions of teosinte (Z. \emph{mays} ssp. \emph{parviglumis}) and maize landraces from the Maize HapMap3 panel as part of the Panzea project (bam files are available at \textcolor{red}{/iplant/home/shared/panzea/hapmap3/bam\_internal/v3\_bams\_bwamem}) \cite{chia2012, lemmon2014,hapmap3}. 
From these data, we removed 4 teosinte individuals that were not ssp. \emph{parviglumis} or appeared as outliers in an initial principal component analysis conducted with the package adegenet \cite{jombart2011} (Figure \ref{sFig:PCA}), leaving 13 teosinte and 23 maize that were used for all subsequent analyses (Table \ref{sTab:list}). We also utilized a single tripsacum (\emph{T. dactyloides}) individual as an outgroup (bam file available at \textcolor{red}{/iplant/home/shared/panzea/hapmap3/bam\_internal/v3\_bams\_bwamem}). \jri{some clarification here. did we use Lemmon's higher depth teosinte? If so, why does Li say we only had 7 high-depth bam files? there should be more}

>>>>

\subsection{Hard sweeps do not shape genome-wide diversity in maize} %JRI

Our findings demonstrate that classic hard selective sweeps have not contributed substantially to genome-wide patterns of diversity in maize, a result we show is robust to concerns about power due to the effects of background selection  \cite{enard2014}. 
Although our approach ignores the potential for hard sweeps in noncoding regions of the genome, a growing body of evidence argues against hard sweeps as the prevalent mode of selection shaping maize variability. 
Among well-characterized domestication loci, only the gene \emph{tga1} shows evidence of a hard sweep on a missense mutation \cite{wang2015}, while several loci are consistent with ``soft sweeps''  from standing variation \cite{studer2011,gallavotti2004role} or multiple mutations \cite{wills2013}. 
Moreover, genome-wide studies of domestication \cite{hufford2012},  local adaptation \cite{Takuno15062015} and modern breeding \cite{beissinger2014} all support the importance of standing variation as the primary source of adaptive variation. 
Soft sweeps are expected to be common when $2N_e\mu_b \ge 1$, where $\mu_b$ is the mutation rate of beneficial alleles with selection coefficient $s_b$ \cite{messer2013population}.  
Assuming a mutation rate of $3 \times 10^{-8}$ \citep{clark2005} and that on the order of $\approx 1-5\%$ of mutations are beneficial \citep{eyre2007distribution}, this implies that soft sweeps should be common in both maize and teosinte for mutational targets $>>10kb$ --- a plausible size for quantitative traits or for regulatory evolution targeting genes with large up- or down-stream control regions \citep[e.g.][]{studer2011}. 
Indeed, many adaptive traits in both maize \citex and teosinte \citex are highly quantitative \jri{cite wallace for maize? weber and/or lauter for teosinte?} and adaptation in both maize \citep{hufford2012} and teosinte \citep{pyhajarvi2013complex} has involved selection on regulatory variation.  \jri{what \% of the windows of top H12 score genome-wide occur outside of genes? is that a number to include here?}

\subsection{Demography of domestication} %JRI
Although a number of authors have investigated the demography of maize domestication \cite{eyre1998, tenaillon2004selection, wright2005}, these  efforts relied on data only from genic regions of the genome and made a number of limiting assumptions about the demographic model.  
We show that diversity within genes has been strongly reduced by the effects of linked selection, such that even synonymous polymorphisms in genes are not representative of diversity at unconstrained sites.  
Furthermore, by utilizing the full joint SFS, we are able to estimate population growth, gene flow, and the strength of the domestication bottleneck without requiring assumptions about its duration.

One surprising result from our model is the estimated timing of domestication at $\approx 15,000$ years before present.
While this appears to conflict with archaeological \cite{piperno2009starch} estimates, we instead argue this estimate reflects the fact that the genetic split between populations likely preceded anatomical changes that can be identified in the archaeological record. 
We also note that our result may also be inflated due to population structure, as our geographically diverse sample of teosinte may include populations diverged from those that gave rise to maize.

The estimated bottleneck of $\approx 5\%$ of the ancestral teosinte population seems low given that maize landraces exhibit $\approx 80\%$ of the diversity of teosinte \citep{hufford2012}, but our model suggests that the effects of the bottleneck on diversity are likely ameliorated by both gene flow and rapid population growth (Figure \ref{fig:bottleneck}).   
Although we estimate that the modern effective size of maize is larger than teosinte, the small size of our sample reduces our power to identify the low frequency alleles most sensitive to rapid population growth \cite{keinan2012}, and our model is unable to incorporate growth faster than exponential.  
Both alternative approaches we employ estimate much a larger modern effective size of maize in the range of $\approx 10^6 - 10^9$, an order of magnitude or more than the current size of teosinte.
Census data suggest these estimates are plausible: there are 47.9 million ha of open-pollinated maize in production \cite{cimmyt1999}, likely planted at a density of $\approx 25,000$ individuals per hectare \cite{baden2001culture}.
Assuming the effective size is only $\approx 0.4\%$ of the census size (i.e. 1 ear for every 1000 male plants), this still implies a modern effective population size of more than four billion.
While these genetic and census estimates are likely inaccurate, all of the evidence points to the fact that the effective size of modern maize is extremely large.

@@

\jri{I would like to see a supplemental figure or maybe just number here on the partial correlation in both teosinte and maize of diversity with recombination after accounting for gene density.  do this in 500kb windows and then in 10kb windows.  we should do the correlation with both pi and singletons.  This let's us compare more explictly with studies like C-D who evaluate overall patterns of linked selection. will be especially cool if we don't see a big difference in big windows but we do in 10kb windows -- siggests C-D looked at wrong scale  }
A consequence of the inefficient purifying selection that maize experienced during its bottleneck is likely that it harbors more weakly deleterious alleles segregating at intermediate frequency than does teosinte. This could be a part of the explanation of why maize inbreds have continued to improve over the past several decades \cite{meghji1984}; if deleterious alleles tend to be recessive and are particularly frequent, they will have ample opportunities to display their phenotypes in inbreds. Our results also demonstrate that recent purifying selection in maize has become much more effective, potentially explaining the ongoing improvement of these inbreds as maize lines are continuously selected. Additionally, the large $N_e$ of modern maize compared to modern teosinte implies that for new mutations, selection will operate much more efficiently in maize. \jri{inbreds stuff is OK, but should be shortened.  we are aiming at a broader audience that doesn't care about maize.  what are more/broader. Chia, Rogers-Melnick, Mezmouk all looked at load. does this have implications for large effect load (i.e. should bneck have purged those?), for heterosis, architecture of quant. traits, etc.}

The absence of evidence for a genome-wide impact of hard sweeps differs markedly from observations in \emph{Drosophila} \cite{sattath2011} and \emph{Capsella} \cite{williamson2014}, but is consistent with data from humans \cite{hernandez2011,pritchard2010genetics}. 
In a related study, \cite{corbett2015} showed that selection tends to have an elevated impact on reducing diversity in species with large population size compared to small. This may explain why maize and humans, both bottlenecked species, show similar patterns while drosophila and capsella behave differently. \textcolor{red}{However, we also do not observe abundant hard sweeps in teosinte, which calls this explanation into question.}

@@

\begin{materials} %JRI

@@

\subsection{Plant materials}
We made use of published sequences from inbred accessions of teosinte (Z. \emph{mays} ssp. \emph{parviglumis}) and maize landraces from the Maize HapMap3 panel as part of the Panzea project  \cite{chia2012, lemmon2014, hapmap3}. 
From these data, we removed 4 teosinte individuals that were not ssp. \emph{parviglumis} or appeared as outliers in an initial principal component analysis conducted with the package adegenet \cite{jombart2011} (Figure \ref{sFig:PCA}), leaving 13 teosinte and 23 maize that were used for all subsequent analyses (Table \ref{sTab:list}). We also utilized a single individual of (\emph{Tripsacum dactyloides}) as an outgroup.  All bam files are available at \textcolor{red}{/iplant/home/shared/panzea/hapmap3/bam\_internal/v3\_bams\_bwamem}). \jri{some clarification here. did we use Lemmon's higher depth teosinte? If so, why does Li say we only had 7 high-depth bam files?  shouldn't there be more?}
